---
title: "QC and pre-processing for amp_phase2_gex project"
subtitle: "Template version 0.2.4"
author: "Baran and McDavid"
date: '`r Sys.Date()`'
params: 
   tenx_root: "scratch/AGG1" # path to raw_feature_bc_matrix tsv mtx triplet
   tenx_h5: NULL # or path to h5. provide only one
   sample_sheet: "all_samples_file.csv" # see ?make_sample_sheet
   qc_output: "refined/qc" #prefix for output
   citeseq_str: 'TotalA' # regular expression matched in `rowData(sce)$Symbol` that identifies "altExp" features. 
                     # See ?partition_citeseq
   batch_var: 'tissue_source'   # if non-null, length-1 character to stratify in cell calling, etc
   covariates: ['dataset', 'sample_ID', 'tissue_source']  #character naming fields in colData to use for plots.  First component should be "most" important value.
   auto_filter: "FALSE" # Should auto-filtering be run
   doublet_detect: ['cxds', 'scDblFinder']
   scDblFinder_args: NULL # or list of arguments, eg `clusters` or `knownDoublets`
   sce_subset: "sce = sce[,sce$detected>500] sce = sce[, sce$subsets_Mito_percent<20]" # character R expression use to subset after QC is run (right before writing data) # should assign to object `sce`
output:
  rmdformats::html_clean:
    code_folding: hide
    toc: yes
    toc_float: yes
---
  
# Set up options
  
```{r options, message = FALSE, warning = FALSE, results = 'hide'}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, autodep=TRUE, message=FALSE, warning=FALSE, cache.lazy = FALSE)
knitr::opts_chunk$set(dev = c('png', 'pdf'))
library(GeneseeSC)
library(dplyr)
library(SingleCellExperiment)
library(scater)
library(DropletUtils)
library(ggplot2)
library(RColorBrewer)
```

This report was compiled with the following parameters:
```{r, cache.whatever=params}
print(params)
```

# Load data
```{r read10x, cache.whatever=list(params$tenx_root, params$citeseq_str)}
if(!is.null(params$tenx_root)){
  agg_raw_data = file.path(params$tenx_root, "outs", "raw_feature_bc_matrix")
  sce = DropletUtils::read10xCounts(agg_raw_data, type="sparse")
} else if(!is.null(params$tenx_h5)){
  sce = DropletUtils::read10xCounts(params$tenx_h5, type="auto")
  assay(sce) = as(assay(sce), 'sparseMatrix')
} else{
  stop("Must provide path to hdf5 format `tenx_h5` or directory with tsv and mtx with `tenx_root`")
}

if (length(params$citeseq_str)>0){
  sce = partition_citeseq(sce, params$citeseq_str)
}
out_size = dim(sce)
```

There were `r out_size[1]` features collected on `r out_size[2]` cells.

# Samples present

```{r, results = 'asis', cache.whatever=list(params$tenx_root, params$sample_sheet, params$citeseq_str)}
sample_sheet = read.csv(params$sample_sheet)
print(knitr::kable(sample_sheet, caption = "Sample Sheet"), "html", table.attr='class="flat-table"')
```

# Link with metadata from sample sheet
```{r expandmeta, results='asis', cache.whatever=list(params$tenx_root, params$sample_sheet, params$batch_var, params$citeseq_str)}
#create metadata for sample based on suffix AGGR adds
sce = expand_metadata(sce, sample_sheet)

count_cells = function(sce, stage){
  colData(sce) %>% as.data.frame() %>% 
  group_by(!!!syms(params$batch_var)) %>% 
  summarize(n_cells = n(), .groups = 'drop') %>% mutate(stage = stage)
} 
qc_table = count_cells(sce, 'droplets')
multiple_batches = length(params$batch_var) > 0 

```

`r if(multiple_batches) {"QC will use batch info"}`
`r if(multiple_batches) {print(params$batch_var)}`

# Cell calling (don't drop)

```{r cell_call, results='asis', cache.whatever=list(params$tenx_root, params$sample_sheet, params$batch_var, params$citeseq_str)}
sce = default_cell_call(sce, params$batch_var, params$covariates[1], drop_empty = FALSE)
sce$called_cell = as.vector(sce$is_cell)
```


## VDJ Discordant Data

```{r}
vdj_dat = list(b = readr::read_csv('extradata/B_no_filter_contigs_cell.csv', guess_max = 1e4),
               t = readr::read_csv('extradata/T_no_filter_contigs_cell.csv', guess_max = 1e4)) %>%
  bind_rows(.id = 'celltype') %>%
  dplyr::select(barcode, sample_ID, tissue_source, sort_group, rawdata_batch, pairing, n_celltypes) 

vdj_dat_uniq = vdj_dat %>%dim(sce)
  group_by(across(barcode:rawdata_batch)) %>% summarize(pairing = any(pairing == 'paired'), n_celltypes = max(n_celltypes), .groups = 'drop') # 300_0171  PBL 1 vs 2 issues.  sample_ID_unique disambiguates these

sce_j = colData(sce) %>% as.data.frame() %>% transmute(barcode = gsub('-[0-9]+', '-1', Barcode),
                                                       sample_ID, tissue_source, sort_group, rawdata_batch, is_cell)
vdj_sce = left_join(sce_j, vdj_dat_uniq)
stopifnot(nrow(sce_j) == nrow(vdj_sce))

sce$has_vdj = case_when(is.na(vdj_sce$pairing) ~ 'No' , vdj_sce$pairing ~ 'Paired chain', TRUE ~ 'Single chain')
sce$vdj_twocelltypes = !is.na(vdj_sce$n_celltypes) & vdj_sce$n_celltypes > 1

# Take no more than 10,000 of each combination of VDJ pairing/presence and called cells
subsample_barcodes = colData(sce) %>% as.data.frame() %>% mutate(idx = seq_len(nrow(.))) %>% group_by(has_vdj, called_cell) %>% slice_sample(n = 1e4)

sce_sub = sce[,subsample_barcodes$idx]
sce_sub$numis = colSums(assay(sce_sub))
sce_sub$ngenes=colSums(assay(sce_sub)>0)
```


```{r}
ggcells(sce_sub, mapping=aes(x=log10(numis), fill=called_cell)) + 
    geom_density() + geom_density(alpha = .3) + facet_wrap(~has_vdj) +ggtitle('Distribution of nUMIs by cell status within BCR/TCR groups')

ggcells(sce_sub, mapping=aes(x=log10(ngenes), fill=called_cell)) + 
    geom_density() + geom_density(alpha = .3) + facet_wrap(~has_vdj) +ggtitle('Distribution of nGenes by cell status within BCR/TCR groups')


ggcells(sce_sub, mapping=aes(x=log10(numis), fill=has_vdj)) + 
    geom_density() + geom_density(alpha = .3) + facet_wrap(~called_cell + tissue_source)+ggtitle('Distribution of nUMIs by BCR/TCR status within cell status')
ggcells(sce_sub, mapping=aes(x=log10(ngenes), fill=has_vdj)) + 
    geom_density() + geom_density(alpha = .3) + facet_wrap(~called_cell + tissue_source)+ggtitle('Distribution of nGenes by BCR/TCR status within cell status')

```

## Define cells (for real)

```{r}
sce = sce[, (colSums(assay(sce))>1000 & sce$tissue_source=='Syn') | (colSums(assay(sce))>1000 & sce$tissue_source=='PBL')]
qc_table = count_cells(sce, 'cells') %>% bind_rows(qc_table)
pre_qc_coldata = colData(sce) %>% as.data.frame()
```

```{r}
table(initially_called_cell = sce$called_cell, vdj = sce$has_vdj)
```

# Default QC and normalization

```{r QCnorm, results='asis', cache.whatever=list(params$tenx_root, params$sample_sheet, params$batch_var, params$citeseq_str, params$auto_filter)}
sce = default_qc(sce, params$batch_var, params$auto_filter, params$covariates) #Doesn't filter anything because auto_filter == FALSE
post_qc_coldata = colData(sce) %>% as.data.frame()
if (params$auto_filter){
qc_table = count_cells(sce, 'qc_pass') %>% bind_rows(qc_table)
}
```
# Doublet detection

## cxds

```{r doub_detect_cxds, results='asis', cache.whatever=list(params$tenx_root, params$sample_sheet, params$batch_var, params$citeseq_str, params$doublet_detect), eval = 'cxds' %in% params$doublet_detect}
sce = doublet_detect_cxds(sce, params$batch_var)
doublet_flag(sce, params$covariates, score_field = 'cxds_score', 'is_singlet_cxds')
```

## scDblFinder

```{r, eval = 'scDblFinder' %in% params$doublet_detect, results='asis', cache.whatever=list(params$tenx_root, params$doublet_detect)}
library(scDblFinder)
# Need version 1.5.16 in Bioc Devel to use known celltypes
#sce = call_intercalate(f = scDblFinder, sce = sce, samples = 'library_id', extra = params$scDblFinder_args)
# 
sce = sce[,sce$dataset!='301_0174_Syn_23'] # breaks scDblFinder
sce = scDblFinder(sce = sce, samples = sce$library_id)
sce$is_singlet_scDbl = (sce$scDblFinder.class == 'singlet')
doublet_flag(sce, params$covariates, score_field = 'scDblFinder.score', 'is_singlet_scDbl')
```

# Normalize

```{r norm, results='asis', cache.whatever=list(params$tenx_root, params$sample_sheet, params$batch_var, params$citeseq_str, params$auto_filter)}
polish_ss = polish_samplesheet_vars(colData(sce)[intersect(names(colData(sce)), names(sample_sheet))])
sce = scater::logNormCounts(sce)
#vars = scater::getVarianceExplained(sce, variables = polish_ss)
#knitr::kable(head(vars))
#scater::plotExplanatoryVariables(vars)
```

```{r}
saveRDS(sce, file = "extradata/PreQCdata.rds")
```


## Optional Subset

```{r subset, eval = !is.null(params$sce_subset), code = params$sce_subset}
```

```{r filter_tag, dev = c('png', 'pdf'), fig.width = 7, fig.height=3.5}
library(dplyr)
library(forcats)
library(ggplot2)

final_qc_cdata = colData(sce) %>% as.data.frame() %>% mutate(post_qc = 'Use')
id_cols = c('library_id', 'Barcode', 'tissue_source', 'sample_ID', 'has_vdj', 'vdj_twocelltypes')
coldata_comp = left_join(pre_qc_coldata[id_cols], final_qc_cdata) %>%
  mutate(post_qc = ifelse(is.na(post_qc), 'Filtered', post_qc),
         doublet = ifelse(is_singlet_scDbl & !vdj_twocelltypes, 'singlet', 'doublet'))

total_syn = final_qc_cdata %>% filter(tissue_source == 'Syn') %>% group_by(sample_ID) %>% 
  summarize(total_cells = n())

coldata_comp = left_join(coldata_comp, total_syn) %>% ungroup() %>%
  mutate(sample_ID = fct_reorder(sample_ID, total_cells))

plt = ggplot(data=coldata_comp, ggplot2::aes(x=sample_ID)) + theme_minimal() + theme(axis.text.x = ggplot2::element_text(angle = 90)) + facet_grid(~tissue_source) + ylab("Cells") + xlab(NULL)

plt + aes(fill = post_qc) +
        ggplot2::geom_bar() + scale_fill_discrete('Cell Quality')

(plt %+% filter(coldata_comp, post_qc == 'Use')) +
        ggplot2::geom_bar() + aes(fill = has_vdj) + scale_fill_discrete('VDJ')
(plt %+% filter(coldata_comp, post_qc == 'Use')) +
        ggplot2::geom_bar()  + aes(fill = doublet) + scale_fill_discrete('')

```

```{r preqc, dev = c('png', 'pdf'), fig.width = 7, fig.height = 3.5}
post_subset_coldata = left_join(colData(sce) %>% as.data.frame, total_syn) %>% ungroup() %>%
  mutate(sample_ID = fct_reorder(sample_ID, total_cells))

(plt %+% post_subset_coldata) + aes(y = detected)+
        ggbeeswarm::geom_quasirandom() + ylab("Genes detected")

(plt %+% post_subset_coldata) + aes(y = log10(sum)) +
        ggbeeswarm::geom_quasirandom() + ylab("log10 UMIs")

(plt %+% post_subset_coldata) + aes(y = subsets_Mito_percent) +
        ggbeeswarm::geom_quasirandom() + ylab("% UMIs Mitochrondial")

```

## Save qc'd data
```{r, cache.whatever=list(params$tenx_root, params$sample_sheet, params$batch_var, params$qc_output, params$citeseq_str, params$auto_filter)}
to_sparse_triplet(sce, params$qc_output)
```
