---
title: "Rates of Somatic Hypermutation for amp_phase2_vdj project"
author: "Wagner, McDavid and Wang"
date: '`r Sys.Date()`'
params:
   all_contigs_ccdb: "refined/B_contigs_ccdb.rds"
   celltype: "B"
   min_cdr3_length: 5
   min_cdr3_dna_length: 15
   cdr3_identity: 0.965
   ident_etc_csv: "refined/02clust_30PC_res05_harmonized__noSCT_Bcells_cdat.csv"
   knitr_cache: FALSE
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, autodep=TRUE, message=FALSE, warning=FALSE, fig.height = 3.5,dev = c('png', 'pdf'))
knitr_cache = params$knitr_cache
knitr::opts_chunk$set(cache = knitr_cache, autodep = knitr_cache, cache.lazy = FALSE)
```

```{r setup, include = FALSE}
library(CellaRepertorium)
library(ggbeeswarm)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(stringr)
library(purrr)
library(knitr)
library(ggraph)
library(igraph)
library(rmdformats)
library(broom.mixed)
library(vegan)
library(ggridges)
library(scales)
#setwd("./B")
```

This file was rendered with the following parameters:
```{r}
print(params)
# Define the category names
ident <- c("B-Naïve(IgD_high)", "B-Naïve(IgD_low)", "B-Naïve(HSPA1B+)", "B-Mito_high", "B-Activated", "B-ABC", "B-Memory", "B-Clonal(LILRA4+)", "Plasmacells", "Plasmablasts", "BT-doublets")
# Define the corresponding colors
colors <- c("#A6CEE3", "#5399C6", "#3E9F32", "#99CD91", "#FDB762", "#CAC848", "#F47676", "#5C0905", "#CCAFCA", "#825D99","grey") 
# Create a named vector
category_colors <- setNames(colors, ident)

tissue <- c("Syn","PBL")
# Define the corresponding colors
tissue_cols <- c("#fcb34d", "#9cb3d1") 
tissue_cols <- setNames(tissue_cols, tissue)
```

# Load filtered contig files
We begin with a dataframe of contigs from the `r params$celltype` cells that have passed through initial quality control.
```{r read_data}
cdb = readRDS(params$all_contigs_ccdb)

if(!is.null(params$ident_etc_csv)){
  ident_etc = readr::read_csv(params$ident_etc_csv) %>% select(has_vdj, discard, keep, Barcode, ident, umap_1, umap_2)
  ident_etc$ident[is.na(ident_etc$ident)] <-"BT-doublets"
  ident_etc$ident <- gsub("\\bPlasma\\b", "Plasmacells", ident_etc$ident)
  cdb$cell_tbl = left_join(cdb$cell_tbl, ident_etc)
}

#Reorder idents to match other plots
cdb$cell_tbl$ident = factor(cdb$cell_tbl$ident, levels = c("B-Naïve(IgD_high)","B-Naïve(IgD_low)","B-Naïve(HSPA1B+)","B-Mito_high",
                                                           "B-Activated","B-ABC","B-Memory","Plasmablasts","Plasmacells","B-Clonal(LILRA4+)"))

#mutate in shm rates:
cdb$contig_tbl = cdb$contig_tbl %>% mutate(v_germ_length = v_germline_end - v_germline_start,
                                             j_germ_length = j_germline_end - j_germline_start,vshm = (1-v_identity)*v_germ_length,
                                             jshm = (1-j_identity)*j_germ_length,shm = vshm + jshm,
                                             len_without_d = v_germ_length + j_germ_length, vj_shm_rate = (shm/len_without_d))
```

```{r sample_Id_recode}
#Some reason ContigCellDB doesn't play well with dplyr::recode, contig_tbl loses basically all recodes with a match when directly updating with cdb$contig_tbl = cdb$contig_tbl %>% mutate(recode())
cell_tbl_recode = cdb$cell_tbl %>% mutate(sample_ID = recode(sample_ID,
                             '300_0150' = 'RA01',
                             '300_0171' = 'RA02',
                             '300_0173' = 'RA03',
                             '300_0174' = 'RA04',
                             '300_0392' = 'RA05',
                             '300_0410' = 'RA06',
                             '300_0414' = 'RA07',
                             '300_0416' = 'RA08',
                             '300_1883' = 'RA09',
                             '300_1930' = 'RA10',
                             '301_0174' = 'RA11',
                             '301_0270' = 'RA12'))

contig_tbl_recode = cdb$contig_tbl %>% mutate(sample_ID = recode(sample_ID,
                             '300_0150' = 'RA01',
                             '300_0171' = 'RA02',
                             '300_0173' = 'RA03',
                             '300_0174' = 'RA04',
                             '300_0392' = 'RA05',
                             '300_0410' = 'RA06',
                             '300_0414' = 'RA07',
                             '300_0416' = 'RA08',
                             '300_1883' = 'RA09',
                             '300_1930' = 'RA10',
                             '301_0174' = 'RA11',
                             '301_0270' = 'RA12'))

cdb@cell_tbl = cell_tbl_recode
cdb@contig_tbl = contig_tbl_recode
```

Initially we start with  `r nrow(cdb)` cells and `r nrow(cdb$contig_tbl)` contigs.  We keep contigs that are 
* full - length 
* productive
* high-confidence
* and with CDR3 sufficiently long, in this report, contigs with CDR3 string length at least `r params$min_cdr3_length`

```{r filtering}
cdb$contig_tbl = dplyr::filter(cdb$contig_tbl, full_length, productive == 'True', high_confidence, chain != 'Multi', 
                               str_length(cdr3) > params$min_cdr3_length, str_length(cdr3_nt) > params$min_cdr3_dna_length)
#Add in cell-level SHM rates
cell_shm = cdb$contig_tbl %>% group_by(!!!syms(cdb$cell_pk)) %>% summarize(shm = mean(vj_shm_rate, na.rm = TRUE))
cdb$cell_tbl = left_join(cdb$cell_tbl, cell_shm, by = cdb$cell_pk)
```

Before filtering
```{r pairing-ident-post, warning=FALSE}
paired_chain = enumerate_pairing(cdb, chain_recode_fun =  if(params$celltype=='B') ig_chain_recode else tcr_chain_recode)
cdb$cell_tbl  = left_join_warn(paired_chain, cdb$cell_tbl, overwrite = TRUE, by = cdb$cell_pk)
```

```{r c_gene_ident}
heavy_chain_ident = cdb %>%  canonicalize_cell(contig_fields = 'c_gene', contig_filter_args = chain=='IGH') %>% `$`('cell_tbl') %>%
  mutate(c_gene = factor(c_gene, levels = c('No VDJ', 'IGHD', 'IGHM', 'IGHG1', 'IGHG2', 'IGHG3', 'IGHG4', 'IGHA1', 'IGHA2')) , c_gene = forcats::fct_explicit_na(c_gene, na_level = 'No VDJ') %>% forcats::fct_rev())

heavy_chain_ident_wide = heavy_chain_ident %>% 
  filter(!is.na(ident))  %>%
  group_by(tissue_source,ident,c_gene, .drop = F) %>% 
  summarize(value = n()) %>% 
  filter(c_gene != "No VDJ") %>%
  ungroup() %>%
  mutate (tissue_source = toupper(tissue_source))%>%
  mutate(new.x =paste(ident,tissue_source,sep = "_"))   %>%
  filter(ident != "B-Clonal(LILRA4+)") %>%
  arrange(c_gene,ident)

chain_pal = c('No VDJ' = 'grey80', 'IGHD' = "#FC9272", 'IGHM'="#DE2D26", 'IGHG1'= "#EFF3FF", 'IGHG2' = "#BDD7E7", 'IGHG3' = "#6BAED6", 'IGHG4' = "#2171B5", 'IGHA1' = "#A1D99B", 'IGHA2' = "#31A354")


Tissue_cols <- ifelse(grepl("SYN", heavy_chain_ident_wide$new.x), "#fcb34d", "#9cb3d1")
isotype_plot = ggplot(heavy_chain_ident_wide %>% filter(!is.na(ident)), aes(x = factor(new.x, levels = unique(new.x)), fill = c_gene, y = value)) + geom_bar(position="fill", stat="identity")+
  theme_classic() + ylab('Proportion')  + scale_fill_manual('', values = chain_pal) + guides(fill=guide_legend(title="IGH Isotype")) +  
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1,colour = Tissue_cols)) 

isotype_plot

readr::write_csv(heavy_chain_ident_wide, "refined/IGH_isotype_plot.csv")
```


# Rates of Somatic Hypermutation (Only for B cells)

# Fit linear mixed effects model using lmer4 to rates of somatic hypermutation with fixed effect for tissue source and random effect for sample
Fixed effect estimates are maximum likelihood estimates, using $\beta \pm 1.96*SE_{\beta}$ for confidence interval

```{r SHM_exploration, eval = as.logical(params$celltype == 'B'), fig.width = 6, fig.height=3.5}
library(broom.mixed)
shm_fitting = function(cdb) {
  #get populations that have at least 1pbl & 1syn cell
  valid_idents_for_fit = cdb$cell_tbl %>% group_by(ident) %>% summarise(n_tissue = length(unique(tissue_source))) %>% filter(!is.na(ident),n_tissue == 2) %>% pull(ident)
  
  #subset cdb
  cell_tbl_valid = cdb$cell_tbl %>% filter(ident %in% valid_idents_for_fit)
  
  #fit lme4::lmer
  shm_fit = cell_tbl_valid %>% group_by(ident) %>% 
  summarize(fit = list(lme4::lmer(shm ~ tissue_source + (1|sample_ID)))) %>%
  rowwise() %>% 
  mutate(td = list(tidy(fit, effects = 'fixed')))
  
  #retreive coefficients
shm_coef = shm_fit %>% select(ident, td) %>% unnest(cols= c(td)) %>% filter(term != '(Intercept)') %>% mutate(p_value = 2*(1-pnorm(abs(statistic))), ci_low = estimate - 1.96*std.error,ci_hi = estimate + 1.96*std.error, tissue_significant = case_when(ci_low > 0 ~ 'Syn', ci_hi < 0 ~ 'PBL', ci_low < 0 & ci_hi > 0 ~ 'NS'))

knitr::kable(shm_coef %>% arrange(desc(statistic)))
return(shm_coef)
}

shm_fit = shm_fitting(cdb = cdb)
shm_fit

label.df = shm_fit %>% select(ident, p_value)  %>% mutate(shm = .2, tissue_source = 'Syn')

mut_plot = ggplot(cdb$cell_tbl %>% filter(!is.na(ident) & ident != "B-Clonal(LILRA4+)"), aes(y = shm * 100, x = tissue_source)) + 
  ylab("V/J SHM %")+ facet_wrap(~ident, ncol = 3) + theme_minimal() + theme(strip.text.x = element_text(size = 6),  strip.background = element_rect(
     color="black", fill="lightblue", linetype="solid")) + scale_color_manual(values = tissue_cols)

shm_plot = mut_plot + 
  geom_quasirandom(aes(color = tissue_source), size = .5)  + 
  stat_summary(color = 'blue', position = position_dodge2(width = .8)) + 
  theme(legend.position = 'none', strip.text.x = element_text(size = 8)) +
  geom_text(data = label.df, aes(x=1.5, y= 22, label = sprintf("p < %1.2g", round(p_value, 3) + 1e-3)))

shm_plot
readr::write_csv(shm_fit, "refined/shm_fit for tissue.csv")
```

Note that for T cells, the "SHM" is just measuring the average number of substitutions vs germline -- not implying hypermutation has taken place in the T cells.

Plot UMIs per cell receptor vs UMAP
Violin plot of this vs identity.

```{r, eval = as.logical(params$celltype == 'B'), fig.width = 8, fig.height=4}
valid_idents_for_fit = cdb$cell_tbl %>% filter(!is.na(ident) & ident != "B-Clonal(LILRA4+)")
mixed.lmer <- lmerTest::lmer(shm~ident +(1|sample_ID) , data = valid_idents_for_fit) 
mixed.lmer_sum <- summary(mixed.lmer)
shm_fit = mixed.lmer_sum$coefficients
shm_fit =as.data.frame(shm_fit)
shm_fit
label.df = data.frame(ident=rownames(shm_fit)[-1],p_value=shm_fit[,"Pr(>|t|)"][-1], shm = .2)
label.df$ident = sub("ident", "", label.df$ident)

mut_plot = ggplot(cdb$cell_tbl %>% filter(!is.na(ident) & ident != "B-Clonal(LILRA4+)"), 
                  aes(y = shm * 100, x = factor(ident, levels = c("B-Naïve(IgD_high)", "B-Naïve(IgD_low)", "B-Naïve(HSPA1B+)", "B-Mito_high", 
                                                                  "B-Activated", "B-ABC", "B-Memory", "Plasmablasts", "Plasmacells")))) + 
  ylab("V/J SHM %")+ theme_minimal() + theme(strip.text.x = element_text(size = 6)) + ylim(0,30)

shm_plot = mut_plot + 
  geom_quasirandom(aes(color = ident), size = .5)  + 
  stat_summary(color = 'blue', position = position_dodge2(width = .8)) + 
  theme_classic(base_size = 16)+ 
  theme(legend.position = 'none', strip.text.x = element_text(size = 8),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  scale_color_manual(values = category_colors) + 
  geom_text(data = label.df, aes(label = sprintf("p < %1.2g", round(p_value, 3) + 1e-3)), size = 4.5, hjust = 0.5, y = 30)+
  labs(x ="", y = paste("V/J SHM %")) 

shm_plot
readr::write_csv(shm_fit, "refined/shm_fit for clusters.csv")
```



